Structure Learning for $\VAR(p)$ Models

### 2019-06-19 paper/paper.tex
Simulation on increasing edge probability, more cleanup on paper.

We proceed first by defining pairwise Granger-causality hypothesis
tests, second a model order selection criteria, and finally the method
for choosing a $P-$value cutoff.

\subsection{Hypothesis Testing}
In performing pairwise checks for Granger-causality $x_j \pwgc x_i$ we
follow the simple scheme of estimating the two linear models via
ordinary least squares:

\begin{align}
  H_0:&\ \widehat{x}_i^{(p_i)}(t) = \sum_{\tau = 1}^{p_i} b_{ii}(\tau)x_i(t - \tau),\\
  H_1:&\ \widehat{x}_i^{(p_i, p_j)}(t) = \sum_{\tau = 1}^{p_i} b_{ii}(\tau)x_i(t - \tau) + \sum_{\tau = 1}^{p_j}b_{ij}(\tau)x_j(t - \tau).
\end{align}

Under the null hypothesis $H_0$ (i.e. that $x_j \npwgc x_i$) we can
form the asymptotically $\chi^2(p_j)$ distributed test statistic
(\hl{TODO: Citation that this is chi2})

\begin{equation}
  \label{eqn:gc_statistics}
  F_{ij}(p_i, p_j) = \frac{T}{p_j}\Big(\frac{\xi_i(p_i)}{\xi_{ij}(p_i, p_j))} - 1\Big),
\end{equation}

@@

\begin{equation*}
  \xi_i(p_i) = \frac{1}{T - p_i}\sum_{t = p_i + 1}^T (x_i(t) - \widehat{x}_i^{(p_i)}(t))^2
\end{equation*}

Given additional assumptions on $x(t)$ (\hl{TODO: What assumptions? --
  this must also depend on choice of $p_i, p_j$}) we have the
convergence (\hl{a.s. or in P?  -- I think in P})

\begin{equation}
  \underset{T \rightarrow \infty}{\text{lim }}F_{ij} =
  \left\{
    \begin{array}{ll}
      0;\ x_j \npwgc x_i\\
      \infty;\ x_j \pwgc x_i
    \end{array}
  \right.
\end{equation}

@@

\subsection{Model Order Selection}
\label{sec:model_order_selection}
There are a variety of methods to choose the filter orders $p_i$ and
$p_j$ \hl{TODO: cite some more}, we will focus in particular on the
Bayesian Information Criteria (BIC).  The BIC is substantially more
conservative than the Akaiake Information Criteria, and since we are
searching for \textit{sparse graphs}, we therefore prefer the BIC. For
the bivariate case we focus for simplicity on $p_i = p_j$ and write
simply\footnote{We use the same $p$ for the bivariate case, but this
  need not be the same as the corresponding univariate estimate}
$x_{ij}^{(p)}$, in which case the BIC, which we seek to
\textit{minimize} over $p$ is given by:

\begin{equation}
  \label{eqn:bic}
  \begin{aligned}
    BIC_{\text{univariate}}(p) &= T \ln \xi_i(p) + p\ln T,\\
    BIC_{\text{bivariate}}(p) &= T \ln \xi_{ij}(p) + 2p\ln T.\\
  \end{aligned}
\end{equation}

We carry this out by a simple direct search on each model order
between $1$ and some prescribed $p_\text{max}$.  In practice it is
sufficient to pick $p_\text{max}$ ad-hoc or via some simple heuristic
(e.g. plotting the sequence $BIC(p)$ over $p$), and we will review in
Section \ref{sec:efficient_model_estimation} an algorithm to carry out
this search efficiently.

% \begin{algorithm}
%   \caption{BIC Model Selection Algorithm}
%   \label{alg:maximize_bic}

%   \SetKwInOut{Input}{input}
%   \SetKwInOut{Output}{output}
%   \SetKwInOut{Initialize}{initialize}
%   \DontPrintSemicolon

%   \BlankLine
%   \Input{$T$ samples of WSS processes $x_i(t), x_j(t)$.  A maximum possible lag $p_{\text{max}}$.}
%   \Output{Model orders $p_i$ and $p$ for models in Hypotheses $H_0$ and $H_1$ respectively.}
%   \Initialize{
%     $BIC_{\text{univariate}} = -\infty$\\
%     $BIC_{\text{bivariate}} = -\infty$\\
%   }

%   \# \texttt{Find $p_i$}\\
%   \For{$p \in \{1, 2, \ldots, p_{\text{max}}\}$}{
%     $\xi_i(p) = $\ \texttt{fit\_univariate}$(x_i(t), p)$ \texttt{ \# Fit model for $H_0$}\\
%     $bic_p = -(T - p - 1) \ln \xi_i(p) - p\ln(T - p - 1)$\\
%     \If{$bic_p > BIC_{\text{univariate}}$}{
%       $BIC_{\text{univariate}} = bic_p$\\
%       \texttt{continue}\\
%     } \Else{
%       $p_i = p - 1$\\
%       \texttt{break}\\
%     }
%   }

%   \BlankLine
%   \# \texttt{Find $p$}\\
%   \For{$p \in \{1, 2, \ldots, p_{\text{max}}\}$}{
%     $\xi_{ij}(p) = $\ \texttt{fit\_biivariate}$(x_i(t), x_j(t), p)$ \texttt{ \# Fit model for $H_1$}\\
%     $bic_p = -(T - p - 1) \ln \xi_{ij}(p) - 2p\ln(T - p - 1)$\\
%     \If{$bic_p > BIC_{\text{bivariate}}$}{
%       $BIC_{\text{biivariate}} = bic_p$\\
%       \texttt{continue}\\
%     }\Else{
%       $p = p - 1$\\
%       \texttt{break}\\
%     }
%   }
%     \Return{$p_i, p$}
% \end{algorithm}

\subsection{Efficient Model Estimation}
\label{sec:efficient_model_estimation}
In practice, the vast majority of computational effort involved in
implementing our estimation algorithm is spent calculating the error
estimates $\xi_i(p_i)$ and $\xi_{ij}(p, p)$.  This requires fitting a
total of $n^2p_{\text{max}}$ autoregressive models, where the most
naive algorithm (e.g. solving a least squares problem for each model)
for this task will consume $O(n^2p_{\text{max}}^4T)$ time, it is
possible to carry out this task in a much more modest
$O(n^2p_{\text{max}}^2 ) + O(n^2p_{\text{max}}T)$ time.

Firstly, we estimate the sequence of Covariance
matrices\footnote{The particular indexing and normalization given in
  equation \ref{eqn:covariance_estimate} is critical to ensure
  $\widehat{R}$ is positive semidefinite.  The estimate can be viewed
  as calculating the covariance sequence of a signal multiplied by a
  rectangular window.}

@@

an $O(n^2p_{\text{max}}T)$ operation.  Secondly, the variance
estimates $\xi_i(p)$ for $p = 1, \ldots, p_{\text{max}}$ can be
evaluated in $O(p_{\text{max}}^2)$ time each by applying the
Levinson-Durbin recursion to $\widehat{R}_{ii}(tau)$, which
effectively estimates a sequence of $AR$ models, producing $\xi_i(p)$
as a side-effect (see
\cite{hayes_statistical_digital_signal_processing} and
\cite{levinson_durbin_recursion}).

Similarly, the variance estimates $\xi_{ij}$ and $\xi_{ji}$ can be
obtained by estimating $\frac{(n + 1)n}{2}$ bivariate AR models, again
in $O(p_{\text{max}}^2)$ via Whittle's generalized LD recursion
\cite{whittle_generalized_levinson_durbin}.

\subsection{Edge Probabilities and Error Rate Controls}
Denote $F_{ij}$ the Granger-causality statistic of equation
\ref{eqn:gc_statistics} with model orders chosen by the methods of
Section \ref{sec:model_order_selection}.  Note that this statistic is
asymptotically $\chi^2(p)$ distributed, and denote by $G$ the
$\chi^2(p)$ cumulative distribution function.  We will define the matrix

@@

The value $\delta$ can be chosen by a variety of methods, in our case
we apply the Benjamini Hochberg criteria \cite{benjamini_hochberg}
\cite{all_of_statistics} to control the false discovery rate of
pairwise edges to a level $\alpha$ (where we generally take
$\alpha = 0.05$).

@@

\section{Empirical Evaluation}
\label{sec:empirical_evaluation}
We have implemented our empirical experiments in Python \cite{scipy},
in particular we leverage the LASSO implementation from
\texttt{sklearn} \cite{sklearn} and the random graph generators from
\texttt{networkx} \cite{networkx}.  We run the same experiment on two
separate graph topologies having $n = 50$ nodes: a strongly causal
graph and a directed acyclic graph.  These are generated respectively
by drawing a random tree and a random Erdos Renyi graph, then
directing edges from lower numbered nodes to higher numbered nodes,
the edge probabilities in the Erdos Renyi graph are tuned to result in
the same number of edges in expectation.

We populate each of the edges (including self loops) with random
linear filters constructed by placing $5$ transfer function poles
(i.e. $p = 5$) uniformly at random in a disc of radius $3 / 4$ (which
guarantees stability for acyclic graphs).  The resulting system is
driven by i.i.d. Gaussian random vectors, each component having random
variance $\sigma_i^2 = 1/2 + r_i$ where $r_i \sim \text{exp}(1/2)$.
We set $p_{\text{max}} = 15$ and collect results for $300$ trials with $T$
varying between $50$ and $5000$.

Results of our Algorithm are compared against the LASSO:

@@

% The pairwise causality function $\mathsf{pwgc}$ of Algorithm
% \ref{alg:Bz_estimate} are obtained via the residuals
% $\mathbf{\epsilon}_{ij}$ of the model
% $\mathbf{x}_i = \mathbf{X}_{ij}\widehat{\beta_{}} +
% \mathbf{\epsilon}_{ij}$ estimated via ordinary least squares, where
% $\mathbf{X}_{ij}$ includes $10$ lagged values of processes $i$ and
% $j$.

% The function $\mathsf{estimate_B}$ in Algorithm \ref{alg:Bz_estimate}
% is carried out via the LASSO \cite{tibshirani2015statistical}, with
% the regularization parameter being chosen to minimize the Bayesian
% information criteria, a more conservative heuristic than the AIC.

@@

  \begin{subfigure}[b]{0.45\textwidth}
    \caption{Random SCG}
    \includegraphics[width=\linewidth]{example_scg.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \caption{Random DAG}
    \includegraphics[width=\linewidth]{example_dag.pdf}
  \end{subfigure}

  \begin{subfigure}[b]{0.45\textwidth}
    \includegraphics[width=\linewidth]{random_scg_simulation.pdf}
    % \missingfigure{}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \includegraphics[width=\linewidth]{random_dag_simulation.pdf}
    % \missingfigure{}
  \end{subfigure}
\end{figure}

In reference to figure \ref{fig:simulation_results} it should not be
overly surprising that our PWGC algorithm performs better than the
LASSO for the case of a strongly causal graph, since in this case the
assumptions which guarantee the correctness of Algorithm
\ref{alg:pwgr_heuristic} hold.  However, the performance is still
markedly superior in the case of a more general DAG.  Finally, we
remark that the PWGC algorithm can be extended to apply to more
general graphs via sequential application.  We will also see in
section \ref{sec:application} that our algorithm can scale effectively
to large problems.

>>>>

In subsection \ref{sec:pairwise_hypothesis_testing} we define pairwise
Granger-causality hypothesis tests, in subsection
\ref{sec:model_order_selection} a model order selection criteria, in
subsection \ref{sec:efficient_model_estimation} an efficient
estimation algorithm, in subsection \ref{sec:error_rate_control} the
method for choosing an hypothesis testing threshold, and finally in
subsection \ref{sec:finite_pwgc} the unified finite sample algorithm.

\subsection{Pairwise Hypothesis Testing}
\label{sec:pairwise_hypothesis_testing}
In performing pairwise checks for Granger-causality $x_j \pwgc x_i$ we
follow the simple scheme of estimating the following two linear models:

\begin{align}
  H_0:&\ \widehat{x}_i^{(p_i)}(t) = \sum_{\tau = 1}^{p_i} b_{ii}(\tau)x_i(t - \tau),\\
  H_1:&\ \widehat{x}_i^{(p_{ij})}(t) = \sum_{\tau = 1}^{p_{ij}} b_{ii}(\tau)x_i(t - \tau) + \sum_{\tau = 1}^{p_{ij}}b_{ij}(\tau)x_j(t - \tau).
\end{align}

Under the null hypothesis $H_0$ (i.e. that $x_j \npwgc x_i$) we can
form the asymptotically $\chi^2(p_{ij})$ distributed test statistic
(\hl{TODO: Citation that this is chi2})

\begin{equation}
  \label{eqn:gc_statistics}
  F_{ij}(p_i, p_{ij}) = \frac{T}{p_{ij}}\Big(\frac{\xi_i(p_i)}{\xi_{ij}(p_{ij})} - 1\Big),
\end{equation}

@@

\begin{equation*}
  \xi_i(p_i) = \frac{1}{T - p_i}\sum_{t = p_i + 1}^T (x_i(t) - \widehat{x}_i^{(p_i)}(t))^2,
\end{equation*}

and similarly for $\xi_{ij}(p_{ij})$.  Under moment conditions $x(t)$
(\hl{TODO: What assumptions? -- this must also depend on choice of
  $p_i, p_j$}) we have the convergence in probability:

\begin{equation}
  F_{ij}(p_i, p_{ij}) \overset{\P}{\rightarrow}
  \left\{
    \begin{array}{ll}
      0;\ x_j \npwgc x_i\\
      \infty;\ x_j \pwgc x_i
    \end{array}
  \right. \text{ as } T \rightarrow \infty.  % the '.' after \right is necessary.
\end{equation}

@@

\subsection{Model Order Selection}
\label{sec:model_order_selection}
There are a variety of methods to choose the filter orders $p_i$ and
$p_j$ (see e.g. \cite{lutkepohl2005new}), but we will focus in
particular on the Bayesian Information Criteria (BIC).  The BIC is
substantially more conservative than the popular alternative Akaiake
Information Criteria (the BIC is also asymptotically consistent), and
since we are searching for \textit{sparse graphs}, we therefore prefer
the BIC, where we seek to \textit{minimize} over $p$:

\begin{equation}
  \label{eqn:bic}
  \begin{aligned}
    BIC_{\text{univariate}}(p) &= \ln\ \xi_i(p) + p\frac{\ln T}{T},\\
    BIC_{\text{bivariate}}(p) &= \ln \det \widehat{\Sigma}_{ij}(p) + 4p\frac{\ln T}{T},\\
  \end{aligned}
\end{equation}

where $\widehat{\Sigma}_{ij}(p)$ is the $2 \times 2$ residual
covariance matrix for the $VAR(p)$ model of $(x_i(t), x_j(t))$.  The
bivariate errors $\xi_{ij}(p)$ and $\xi_{ji}(p)$ are the diagonal
entries of $\widehat{\Sigma}_{ij}(p)$.

We carry this out by a simple direct search on each model order
between $0$ and some prescribed $p_\text{max}$.  In practice it is
sufficient to pick $p_\text{max}$ ad-hoc or via some simple heuristic
e.g. plotting the sequence $BIC(p)$ over $p$, though it is not
technically possible to guarantee that the optimal $p$ is chosen
(since there can in general be arbitrarily long lags from one variable
to another).

\subsection{Efficient Model Estimation}
\label{sec:efficient_model_estimation}
In practice, the vast majority of computational effort involved in
implementing our estimation algorithm is spent calculating the error
estimates $\xi_i(p_i)$ and $\xi_{ij}(p_{ij})$.  This requires fitting
a total of $n^2p_{\text{max}}$ autoregressive models, where the most
naive algorithm (e.g. solving a least squares problem for each model)
for this task will consume $O(n^2p_{\text{max}}^4T)$ time, it is
possible to carry out this task in a much more modest
$O(n^2p_{\text{max}}^2 ) + O(n^2p_{\text{max}}T)$ time via the
autocorrelation method which deploys the following autocovariance
estimates in the Yule-Walker equations:\footnote{The particular
  indexing and normalization given in equation
  \ref{eqn:covariance_estimate} is critical to ensure $\widehat{R}$ is
  positive semidefinite.  The estimate can be viewed as calculating
  the covariance sequence of a signal multiplied by a rectangular
  window.}

@@

These covariance estimates constitute the $O(n^2p_{\text{max}}T)$
operation.  Given these particular estimates, the variances $\xi_i(p)$
for $p = 1, \ldots, p_{\text{max}}$ can be evaluated in
$O(p_{\text{max}}^2)$ time each by applying the Levinson-Durbin
recursion to $\widehat{R}_{ii}(\tau)$, which effectively estimates a
sequence of $AR$ models, producing $\xi_i(p)$ as a side-effect (see
\cite{hayes_statistical_digital_signal_processing} and
\cite{levinson_durbin_recursion}).

Similarly, the variance estimates $\widehat{\Sigma}_{ij}(p)$ (which
include $\xi_{ij}$ and $\xi_{ji}$) can be obtained by estimating
$\frac{(n + 1)n}{2}$ bivariate AR models, again in
$O(p_{\text{max}}^2)$ time via Whittle's generalized Levinson-Durbin
recursion\footnote{We have made use of standalone tailor made
  implementations of these algorithms, available at
  \textsf{github.com/RJTK/Levinson-Durbin-Recursion}.}
\cite{whittle_generalized_levinson_durbin}.

\subsection{Edge Probabilities and Error Rate Controls}
\label{sec:error_rate_control}
Denote $F_{ij}$ the Granger-causality statistic of equation
\ref{eqn:gc_statistics} with model orders chosen by the methods of
Section \ref{sec:model_order_selection}.  Note that this statistic is
asymptotically $\chi^2(p_{ij})$ distributed, and denote by $G$ the
cumulative distribution function thereof.  We will define the matrix

@@

The value $\delta$ can be chosen by a variety of methods, in our case
we apply the Benjamini Hochberg criteria \cite{benjamini_hochberg}
\cite{all_of_statistics} to control the false discovery rate of
pairwise edges to a level $\alpha$ (where we generally take
$\alpha = 0.05$).  Applying the extremely simple Bonferroni method is
overly conservative for this application.

\subsection{Finite Sample Recovery Algorithm}
\label{sec:finite_pwgc}

    \BlankLine
    \caption{Finite Sample Pairwise Graph Recovery (PWGC)}
    \label{alg:finite_pwgc}

\end{algorithm}

@@

\section{Empirical Evaluation}
\label{sec:empirical_evaluation}
We have implemented our empirical experiments in Python \cite{scipy},
in particular we leverage the LASSO implementation from
\texttt{sklearn} \cite{sklearn} and the random graph generators from
\texttt{networkx} \cite{networkx}.  We run experiments using two
separate graph topologies having $n$ nodes: a strongly causal graph
(SCG) and a directed acyclic graph (DAG).  These are generated
respectively by drawing a random tree and a random Erdos Renyi graph
(with edge probability $q = \frac{2}{n}$ resulting in approximately
the same number of edges for the SCG as for the DAG), then creating a
directed graph by directing edges from lower numbered nodes to higher
numbered nodes.

We populate each of the edges (including self loops) with random
linear filters constructed by placing $5$ transfer function poles
(i.e. $p = 5$) uniformly at random in a disc of radius $3 / 4$ (which
guarantees stability for acyclic graphs).  The resulting system is
driven by i.i.d. Gaussian random noise, each component having random
variance $\sigma_i^2 = 1/2 + r_i$ where $r_i \sim \text{exp}(1/2)$.
We set $p_{\text{max}} = 15$.  Results and representitive graphs are
collected in Figure \ref{fig:simulation_results}.

We compare our results against the LASSO as a baseline, estimating the
filter $\B(z)$ that minimizes $\xi_{\text{LASSO}}$ in the following:

@@

@@

  {\scriptsize Comparison of PWGC and LASSO for $VAR(p)$ model
    estimation.  The MCC is measured on the left axis in blue with
    circle markers, and the relative log mean-squared prediction error
    $\frac{\ln\ \tr \widehat{\Sigma}_v}{\ln\ \tr \Sigma_v}$ on the
    right axis in red with triangle markers.  We use $q$ to denote the
    edge probability in a random Erdos-Reny graph (modified to form a
    DAG).  Solid lines indicate ad-hoc regression estimates for
    visualization purposes only.

    Matthew's Correlation Coefficient (MCC)
    \cite{matthews1975comparison} synthesizes the confusion matrix
    into a single score appropriate for unbalanced labels and is
    calibrated to fall into the range $[-1, 1]$ with $1$ being perfect
    performance, $0$ being the performance of random guessing, and
    $-1$ being perfectly opposed.
  }

  \begin{subfigure}[b]{0.3\textwidth}
    \caption{Random SCG}
    \includegraphics[width=\linewidth]{example_scg.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\textwidth}
    \caption{Random DAG $(q = \frac{2}{n})$}
    \includegraphics[width=\linewidth]{example_dag.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\textwidth}
    \caption{Random DAG $(q = \frac{4}{n})$}
    \includegraphics[width=\linewidth]{example_dag_q.pdf}
  \end{subfigure}

  \begin{subfigure}[b]{0.45\textwidth}
    \caption{PWGC vs LASSO (SCG)}
    \label{fig:simulation_results_comparison1}
    \includegraphics[width=\linewidth]{random_scg_simulation.pdf}
    % \missingfigure{}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \caption{PWGC vs LASSO (DAG, $q = \frac{2}{n}$)}
    \label{fig:simulation_results_comparison2}
    \includegraphics[width=\linewidth]{random_dag_simulation.pdf}
    % \missingfigure{}
  \end{subfigure}

  \begin{subfigure}[b]{0.45\textwidth}
    \caption{Fixed $T$, increasing $n$ (SCG)}
    \label{fig:simulation_results_scaling}
    \includegraphics[width=\linewidth]{increasing_n_simulation.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \caption{Fixed $T, n$, increasing edges $q$ (DAG)}
    \label{fig:simulation_results_dense}
    \includegraphics[width=\linewidth]{dag_increasing_q_simulation.pdf}
  \end{subfigure}
\end{figure}

In reference to figure \ref{fig:simulation_results_comparison1} it
should not be overly surprising that our PWGC algorithm performs
better than the LASSO for the case of a strongly causal graph, since
in this case the assumptions which guarantee the correctness of
Algorithm \ref{alg:pwgr_heuristic} hold.  However, the performance is
still markedly superior in the case of a more general DAG.

In regards scalability, we have observed that performing the $O(n^2)$
pairwise Granger-causality calculations consumes the vast majority
($> 90\%$) of the computation time.  Since this step is trivially
parallelizable, our algorithm also scales well with multiple cores or
multiple machines.  Figure \ref{fig:simulation_results_scaling} is a
demonstrating of this scalability, where we are able to estimate
graphs having over $1500$ nodes (over $2.25 \times 10 ^6$ possible
edges) using only $T = 500$ data points, granted, an SCG on this many
nodes is extremely sparse.  Application to denser graphs is explored
in Figure \ref{fig:simulation_results_dense}, where $n$ and $T$ are
held fixed while increasing the edge probability $q$, thereby creating
denser graphs.  For larger values of $q$ in this plot there is a
comparable number of edges as there are data samples $T$.
